Relational Parametricity for Algebraically Indexed Types

### 2012-06-07 paper/geometry-twocolumn.tex
Substantial changes to the motivating examples section

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of real numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ may be called upon to represent both
\emph{points} -- offsets from some origin -- and \emph{vectors} --
offsets in their own right. The notions of point and vector are very
different despite their common representation. By representing a point
concretely as an offset from an origin, we inadvertently fix that
origin as privileged.

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
follow the general principle that our programs are invariant with
respect to changes in the choice of origin. On the other hand, vectors
are not intended to be regarded as invariant under change of choice of
origin. Vectors represent offsets, or movements, and the vector
$(0,0)$ always represents the zero offset.
% However, they are intended to be invariant under change of basis: the
% vector $(2,3)$, with respect to the basis $(1,0),(0,1)$, and the
% vector $(1,1)$, with respect to the basis $(2,0),(0,3)$, should be
% regarded as the same vector. Invariance under change of basis also
% applies to the representation of points. We therefore refine the
% principle of invariance under change of origin for points to also
% include invariant under change of basis for points and vectors.

% A particular choice of origin and basis is called a \emph{frame}
% (\cite{gallier11geometric}, Chapter 2). In general, the technique of
% maintaining geometric invariants such as frame invariance has been
% referred to as \emph{coordinate-free geometry}: we should aim to
% think in terms of abstract concepts such as points and vectors,
% rather than their actual representation as tuples of numbers.

Invariance under change of representation immediately recalls
Reynolds' fable concerning two professors teaching the theory of
complex numbers \cite{reynolds83types}. One professor chooses
rectangular coordinates ($x + iy$), while the other chooses polar
coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
learning the basic operations on complex numbers in the two
representations, the two classes can interact because the theory of
complex numbers is invariant with respect to the choice of
representation. Reynolds formalises the idea of invariance with
respect to changes of representation by using the notion of
preservation of relations. If a binary relation $R$ relates two
representations of a concept---for example the rectangular and polar
representations of the same complex number are related---then a
program that manipulates complex numbers at a level of abstraction
above their specific representation should preserve the relation
$R$. For example, if $f$ be a program taking complex numbers to
complex numbers that is intended to respect invariance under change of
representation. Then if $c$ is a complex number in rectangular form,
and $c'$ is a complex number in polar form such that $R$ relates them,
which we write as $(c,c') \in R$, then it should be the case that
$(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to defining invariance
under changes of representation to the geometric setting. Following
Reynolds, we express the fact that two tuples in $\mathbb{R}^2$ are
representatives of the same point under change of origin by defining a
family of binary relations over $\mathbb{R}^2$. Our family of binary
relations is parameterised by a particular choice of origin, and by
quantifying over all choices of origin we will be able to ensure that
our programs are invariant with respect to the particular choice of
origin. We define a $\mathbb{R}^2$-parameterised family of binary
relations $\{ R_{\vec{o}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
\}_{\vec{o} \in \mathbb{R}^2}$ as follows:
\begin{displaymath}
  R_{\vec{o}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{o} \}
\end{displaymath}

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points with respect to the same
origin, but that the function itself should be invariant with respect
to the choice of origin. We use Reynolds' idea of preservation of
relations to formalise this: for any $\vec{o} \in \mathbb{R}^2$
(i.e.~any choice of origin), the function $f$ should satisfy the
following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{o}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{o}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{o}}
\end{equation}
Unfolding the definition of $R_{\vec{o}}$, this statement is
equivalent to the following statement expressing invariance of $f$
with respect to changes of origin: for all $\vec{o} \in \mathbb{R}^2$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{o},\vec{v_2} + \vec{o}) = f(\vec{v_1},\vec{v_2}) + \vec{o}.
\end{equation}
So Reynolds' preservation of relations, when instantiated with the
family of relations $R$, yields exactly the geometric property of
invariance under change of origin.

@@

Reynolds further insight was to see how preservation of relations by
the denotational interpretations of programs can automatically be
established by following a typing discipline. In Reynolds' case, the
typing discipline was the polymorphic $\lambda$-calculus where new
types can be constructed by universal quantification over all
types. Universal type quantification is commonly written $\forall
\alpha. A$, which denotes a type that may be instantiated with any
other type $B$ by replacing the type variable $\alpha$ with $B$ in
$A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all
choices of origin and used the choice to select a relation from the
family $\{R_{\vec{o}}\}$. Driven by this observation, we introduce
quantification over choice of origin into the language of types.  We
use the notation $\forall t \mathord: T_2. A$ for quantification over
all 2-dimensional translations (i.e.~choices of origin) $t$. We refer
to $T_2$ as the \emph{sort} of $t$.

% As we proceed, we will introduce the sorts representing geometric
% groups that we will be using in addition to $\GA(2)$ and $\GL(2)$.

The sort $T_2$ represents an abelian group, so we can combine its
elements using the usual group operations. We write operations
additively, using $e_1 + e_2$ for the group operation, $-e$ for
inverse and $0$ for the unit.  We also regard expressions built from
variables and the group operations up to the group axioms. So, for
example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\mathsf{Unit}$ and tuple types $A \times B$. We also assume a
primitive type $\mathsf{real}$, used to represent scalars. As we noted
at the start of the previous section, the central data structure in
geometric applications is the tuple of real numbers for representing
points and vectors. However, we cannot simply express this as the type
$\mathsf{real} \times \mathsf{real}$ because it will not have the
correct relational interpretation. (Two elements of type
$\mathsf{real}$ will be related if and only if they are equal, and
hence two elements of $\mathsf{real} \times \mathsf{real}$ will be
related if and only if they are equal, by Reynolds' definition of the
relational interpretation of tuple types.) Therefore, we introduce a
new type parameterised by expressions $e$ of sort $T_2$ to represent
points (and later, vectors): $\mathsf{vec}\langle e \rangle$. We have
used the name $\mathsf{vec}$, even though we have taken pains to
separate geometric points and vectors, to recall the computer science
notion of vector as a sequence of values of homogeneous type which has
a known length (in this case two).

% FIXME: Denotational semantics of $\mathsf{vec}$? and transformation
% quantification?

Equipped with quantification over the group $T_2$ of changes of
origin, and the parameterised types $\mathsf{vec}\langle \cdot
\rangle$, we can now express origin invariance properties as types. In
the previous section, we discussed functions $f : \mathbb{R}^2 \times
\mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \times \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle
\end{displaymath}
FIXME: forward references...

@@

As we stated in the previous section, the primitive operation of
affine geometry is affine combination of points: $\lambda_1\vec{v_1} +
\lambda_2\vec{v_2}$, where $\lambda_1 + \lambda_2 = 1$. Geometrically,
affine combination can be interpreted as interpolation between the
points represented by $\vec{v_1}$ and $\vec{v_2}$. We add affine
combination of points to our calculus, with the following typing and
intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affineCombination} :\forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \\
    \sem{\mathrm{affineCombination}}\ (x_1,y_1)\ r\ (x_2,y_2) = \\
    \hspace{3cm}((1-r)x_1 + tx_2, (1-r)y_1 + ty_2)
  \end{array}
\end{displaymath}
By defining the primitive function $\mathrm{affineCombination}$ to
take a single $\mathsf{real}$ parameter $t$, we can easily ensure that
we are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination. The intended denotation
in the second line does not take any argument corresponding to the
universally quantified $t$ in the type. Type-level variables are
erased at runtime.

% In Linear Algebra notation, we could have written the right-hand side
% as $(1-t)\vec{v_1} + t\vec{v_2}$, where $\vec{v_1} = (x_1,y_1)$ and
% $\vec{v_2} = (x_2,y_2)$.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{real} \to \mathsf{vec}\langle t \rangle \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad
      \begin{array}{l@{\hspace{-0.002cm}}l}
        \mathrm{affineCombination}\ [F]\ & (\mathrm{affineCombination}\ [F]\ p_0\ s\ p_1) \\
        & s \\
        & (\mathrm{affineCombination}\ [F]\ p_1\ s\ p_2)
      \end{array}
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin
\end{example}

Affine combination only provides us with an operation on points. We
are also interested in operations on the vectors representing offsets
between points. We now examine the correct types to assign to the
vector space operations of addition of vectors, negation of vectors,
multiplication by a scalar and the zero vector. The typings of these
operations will make use of the abelian group structure of the sort of
translations $T_2$.

Vector operations are not invariant under change of origin. The
obvious type assignment for vector addition is therefore:
\begin{displaymath}
  (+) : \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2 \rangle \to \mathsf{vec}\langle t_1 + t_2 \rangle
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle -t \rangle
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:T_2.\ \mathsf{vec}\langle t_1\rangle \to \mathsf{vec}\langle t_2\rangle \to \mathsf{vec}\langle t_1 - t_2 \rangle \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\mathsf{vec}\langle t \rangle$---we can compute their offset, which
is a vector expressed with respect to the null change of origin. The
offset operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The type of the vector addition operation can be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:T_2.\ \mathsf{vec}\langle t \rangle \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle t \rangle \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations%\footnote{but they could...}
. The zero vector has the type:
\begin{displaymath}
  \mathrm{0} : \mathsf{vec}\langle 0 \rangle
\end{displaymath}
Intuitively, there cannot be a zero vector invariant under any change
of origin, because this would entail picking a privileged origin. For
multiplication of a vector by a scalar, we assign the following type,
which is again not invariant under change of origin.
\begin{displaymath}
  (*) : \mathsf{real} \to \mathsf{vec}\langle 0 \rangle \to \mathsf{vec}\langle 0 \rangle
\end{displaymath}

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:T_2.\ \underbrace{\mathsf{vec}\langle t \rangle \to ... \mathsf{vec}\langle t \rangle}_{n+1\textrm{ times}} \to \mathsf{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $T_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n}$ are isomorphic to the corresponding types
  $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\mathsf{vec}\langle 0 \rangle \to ... \mathsf{vec}\langle 0 \rangle}_{n\textrm{ times}} \to \mathsf{real}
  \end{displaymath}
  We demonstrate this isomorphism formally in \autoref{sec:FIXME}.

@@

\begin{example}
  An inexpressible type
  \begin{displaymath}
    \forall t \mathord: T_2.\ \vec{t + t} \to \vec{t}
  \end{displaymath}
  FIXME: explain that it is possible to show that this type is
  uninhabited with the given operations. We show this in general 
\end{example}

\subsection{Generalising Origin Invariance}
\label{sec:motivation-generalising}

\begin{enumerate}
\item Basis invariance
\item Orthogonal transform invariance for dot product
\item Rotational invariance (the type of the cross-product)
\item Scale invariance (units of measure)
\end{enumerate}

In some cases, we can use the invariance as a programming tool
(e.g. type isomorphisms), but at others we can only use it from the
outside.

>>>>

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of real numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ may be called upon to represent both
\emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right. Points and vectors are
very different despite their common representation.

Distinguishing between points and vectors is the key feature of
\emph{affine geometry} (see, for example, Gallier's book
\cite{gallier11geometric}, Chapter 2). Traditionally in computational
geometry libraries, the distinction between points and vectors has
either been left to the programmer to maintain by themselves, or has
been enforced through the use of separate abstract types for points
and vectors (one might say that the standard mathematical formulation
of affine spaces uses the abstract types approach). Here, we
investigate a more sophisticated approach using types indexed by
change of origin transformations, and think of the difference between
points and vectors as a change of data representation.

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
ensure that our programs are invariant with respect to changes in the
choice of origin. On the other hand, vectors are not intended to be
regarded as invariant under change of origin. Vectors represent
offsets, or movements, and the vector $(0,0)$ always represents the
zero offset.

Invariance under change of representation immediately recalls
Reynolds' fable concerning two professors teaching the theory of
complex numbers \cite{reynolds83types}. One professor chooses
rectangular coordinates ($x + iy$), while the other chooses polar
coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
learning the basic operations on complex numbers in the two
representations, the two classes can interact because the theory of
complex numbers is invariant with respect to the choice of
representation. Reynolds formalises the idea of invariance with
respect to changes of representation as preservation of relations. If
a binary relation $R$ relates two representations of a concept---for
example the rectangular and polar representations of the same complex
number are related---then a program that manipulates complex numbers
at a level of abstraction above their specific representation should
preserve the relation $R$. For example, let $f$ be a program taking
complex numbers to complex numbers that is intended to respect
invariance under change of representation. Then if $c$ is a complex
number in rectangular form, and $c'$ is a complex number in polar form
such that $R$ relates them, which we write as $(c,c') \in R$, then it
should be the case that $(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to the geometric
setting. We define a family of binary relations on $\mathbb{R}^2$
which is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group under
addition. To emphasise the group structure, and the intended meaning
as translations, we write the collection of all change of origin
offsets as $\Transl_2$. By quantifying over all changes of origin we
will be able to ensure that our programs are invariant with respect to
the particular choice of origin. The members of the
$\Transl_2$-indexed family of binary relations $\{ R_{\vec{t}}
\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl_2}$
are defined as follows:
\begin{displaymath}
  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
\end{displaymath}

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points with respect to the same
origin, but that the function itself should be invariant with respect
to the choice of origin. We use Reynolds' idea of preservation of
relations to formalise this: for any $\vec{t} \in \Transl_2$ (i.e.~any
change of origin), the function $f$ should satisfy the following
statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$, this statement is
equivalent to the following statement expressing invariance of $f$
with respect to changes of origin: for all $\vec{t} \in \Transl_2$:
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} + \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
So Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

@@

Reynolds' further insight was to see how preservation of relations by
the denotational interpretations of programs can automatically be
established by following a typing discipline. In Reynolds' case, the
typing discipline was the polymorphic $\lambda$-calculus where new
types can be constructed by universal quantification over all
types. Universal type quantification is commonly written $\forall
\alpha. A$, which denotes a type that may be instantiated with any
other type $B$ by replacing the type variable $\alpha$ with $B$ in
$A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all
changes of origin and used the choice to select a relation from the
family $\{R_{\vec{t}}\}$. Driven by this observation, we introduce
quantification over changes of origin into the language of types.  We
use the notation $\forall t \mathord: \SynTransl_2. A$ for
quantification over all 2-dimensional translations (i.e.~choices of
origin) $t$. We refer to $\SynTransl_2$ as the \emph{sort} of
$t$. Note that we have used a different font to distinguish the
semantic group $\Transl_2$ from the syntactic sort $\SynTransl_2$.

The sort $\SynTransl_2$ represents an abelian group, so we can combine
its elements using the usual group operations. We write operations
additively, using $e_1 + e_2$ for the group operation, $-e$ for
inverse and $0$ for the unit.  We also regard expressions built from
variables and the group operations up to the abelian group axioms. For
example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
primitive type $\tyPrimNm{real}$, used to represent scalars. As we
noted at the start of the previous section, the central data structure
in geometric applications is the tuple of real numbers for
representing points and vectors. However, we cannot simply express
this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
will not have the correct relational interpretation. (Two elements of
type $\tyPrimNm{real}$ will be related if and only if they are equal,
and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
will be related if and only if they are equal, by Reynolds' definition
of the relational interpretation of tuple types.) Therefore, we
introduce a new type parameterised by expressions $e$ of sort
$\SynTransl_2$ to represent points \emph{and} vectors:
$\tyPrim{vec}{e}$. We have used the name $\tyPrimNm{vec}$, even though
we have taken pains to separate geometric points and vectors, to
recall the computer science notion of vector as a sequence of values
of homogeneous type which has a known length (in this case two).

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl_2$
into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{t}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of the universal quantification
$\forall t\mathord:\SynTransl_2.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
depends on which (potentially non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables to the group $\Transl_2$, and extension as just adding an
extra component to the tuple. The general index-erasure semantics will
be given in \autoref{sec:index-erasure-semantics}. The full relational
interpretation is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

Equipped with quantification over the group $\SynTransl_2$ of changes
of origin, and the indexed types $\tyPrim{vec}{\cdot}$, we can now
express origin invariance properties as types. In the previous
section, we discussed functions $f : \mathbb{R}^2 \times \mathbb{R}^2
\to \mathbb{R}^2$ that have the property that they preserve all
changes of origin. We can now express this property as a type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above (and the standard relational interpretations for
tuple and function types), we obtain
\statementref{eq:f-invariant-frame}.

@@

Invariance under change of origin is the key feature of affine
geometry, and the central operation of affine geometry is the affine
combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
can be interpreted as interpolation between the points represented by
$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
our calculus, with the following typing and intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-t)\vec{v_1} + t\vec{v_2}
  \end{array}
\end{displaymath}
It can be verified by hand that the intended denotation
$\sem{\mathrm{affComb}}$ has the invariance property specified by its
type.  By defining the primitive function $\mathrm{affComb}$ to take a
single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad \mathrm{affComb}\ [F]\ (\mathrm{affComb}\ [F]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [F]\ p_1\ s\ p_2) \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin.
\end{example}

Affine combination only provides us with an operation on points. We
are also interested in operations on the vectors representing offsets
between points. We now examine the correct types to assign to the
vector space operations of addition of vectors, negation of vectors,
multiplication by a scalar and the zero vector. The typings of these
operations will make use of the abelian group structure of change of
origin translations.

Vector operations are not invariant under change of origin. The
obvious type for vector addition is therefore:
\begin{displaymath}
  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The type of the vector addition operation can be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}\\
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

% \fixme{If we had dependent types, could the affComb operation be
%   derived from an appropriately dependently typed scaling operation?}

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl_2.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $\SynTransl_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

  The types $\tau_{n}$ are isomorphic to the corresponding types
  $\sigma_{n}$:
  \begin{displaymath}
    \sigma_{n} \isDefinedAs \underbrace{\tyPrim{vec}{0} \to ... \tyPrim{vec}{0}}_{n\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  We demonstrate this isomorphism formally in
  \autoref{sec:instantiations}, in a more general setting of types
  indexed by abelian groups.

@@

\begin{example}
  An inexpressible type
  \begin{displaymath}
    \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
  \end{displaymath}
  FIXME: explain that it is possible to show that this type is
  uninhabited with the given operations.
\end{example}

\subsection{Change of Basis Invariance}
\label{sec:motivation-generalising}

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations. For origin
invariance, we quantified over members of the group $\SynTransl_2$ of
$2$-dimensional translations. Members of this group represent changes
of origin. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL_2$.

\fixme{new relational interpretation of vectors}

\paragraph{Affine Geometry} The combination of a change of basis
transformation and a change of origin transformation is known as an
\emph{affine transformation}. A way of describing affine geometry is
as the collection of properties that are invariant under invertible
affine transformations.
